Compute the distance between `POINT(10 -90)` and `POINT(50 -90)`, assuming (i) these are coordinates in a Cartesian space, and (ii) these are geodetic coordinates. What are the units of the result?
```{r}
library(sf)
st_as_sfc(c("POINT(10 -90)", "POINT(50 -90)")) |> 
		st_distance()
st_as_sfc(c("POINT(10 -90)", "POINT(50 -90)"), crs = st_crs('OGC:CRS84')) |> 
		st_distance()
```

Load the `nc` dataset into your session (e.g. using `library(sf); demo(nc)`) and convert it into a `stars` object using (i) `st_as_stars()`, (ii) `st_rasterize()`, (iii) `st_interpolate_aw()`
```{r}
demo(nc, echo = FALSE, ask = FALSE)
library(stars)
(r1 <- st_as_stars(nc))
(r2 <- st_rasterize(nc))
# gr = st_make_grid(nc)
gr = st_as_stars(st_bbox(nc), nx = 10, ny = 10)
(r3 <- st_interpolate_aw(nc["SID74"], gr, extensive = TRUE))
```

Load the `L7_ETMs` dataset into your session (e.g. using `library(stars); L7_ETMs = st_as_stars(L7_ETMs)`), and convert the object to an `sf` object (i) using `st_as_sf()`, (ii) using `st_as_sf(..., as_points = TRUE)`, and explain the differences (also plot the resulting `sf` objects). Randomly sample 100 points from the bounding box of `L7_ETMs`, and extract the image values at these points using `st_extract()`, and convert the result into an `sf` object.
```{r}
L7_ETMs = st_as_stars(L7_ETMs)
(s1 <- st_as_sf(L7_ETMs))
(s2 <- st_as_sf(L7_ETMs, as_points = TRUE))
plot(s1, border = NA) # without border=NA, you may only see polygon borders
plot(s2)
pts = st_sample(st_bbox(L7_ETMs), 100)
(e1 <- st_extract(L7_ETMs, pts) |> st_as_sf())
(e2 <- st_intersection(s1, pts))
all.equal(e1, e2, check.attributes = FALSE)
(e3 <- st_intersection(s2, pts))
```


From the point pattern shown in section 1.3, download the GeoPackage, and read into R
```{r}
library(sf)
library(rnaturalearth)
w = read_sf("Windkraftanlagen_DE_5521464311407255742.gpkg")
```

Read the boundary of Germany using `rnaturalearth::ne_countries(scale = "larger", country = "Germany")
```{r}
g = ne_countries(scale = "large", country = "Germany") |> st_transform(st_crs(w))
```

Create a plot showing both the observation window and 
```{r}
plot(st_geometry(g))
plot(st_geometry(w), add = TRUE, pch = 3, cex = .2)
```

Do all observations fall inside the observation window?
```{r}
st_disjoint(w, g) |> lengths() |> sum()
```

Create a ppp object from the points and the window
```{r}
library(spatstat)
c(st_geometry(g), st_geometry(w)) |> as.ppp() -> pp
# or, with marks:
pp <- as.ppp(w[g, ], W = as.owin(g))
```

Create a density map of the wind turbines, with the turbines added
```{r}
plot(density(pp))
plot(pp, add = TRUE, pch = 3, cex = .2, col = 'green')
```

CSR?
```{r}
(q = quadrat.test(pp))
plot(q)
```

density:
```{r}
d <- density(pp)
# verify density:
mean(d) * st_area(g) |> units::drop_units()
nobjects(pp)
```

# interactions?
```{r}
Kest(pp) |> plot()
Lest(pp) |> plot()
Linhom(pp) |> plot()
```
These two take a rather long time: try it by yourself:
```{r eval=FALSE}
envelope(pp, Lest) |> plot()
envelope(pp, Linhom) |> plot()
```
