## Exercises

Compute the variogram cloud of NO2 using `variogram()` and  argument
`cloud = TRUE`.  (a) How does the resulting object differ from the
"regular" variogram (use the `head` command on both objects); 
(b) what do the "left" and "right" fields refer
to? (c) when we plot the resulting variogram cloud object, does it still indicate
spatial correlation?

```{r}
library(sf)
no2 <- read.csv(system.file("external/no2.csv",
    package = "gstat"))
crs <- st_crs("EPSG:32632") # a csv doesn't carry a CRS!
st_as_sf(no2, crs = "OGC:CRS84", coords =
    c("station_longitude_deg", "station_latitude_deg")) |>
    st_transform(crs) -> no2.sf
library(ggplot2)
# plot(st_geometry(no2.sf))
"https://github.com/edzer/sdsr/raw/main/data/de_nuts1.gpkg" |>
  read_sf() |>
  st_transform(crs) -> de
library(gstat)
variogram(NO2~1, no2.sf, cloud = TRUE, cutoff = 350000) |> 
		plot()
```

2. Compute the variogram of NO2 as above, and change the arguments `cutoff`
and `width` into very large or small values. What do they do?

```{r}
variogram(NO2~1, no2.sf, width=50, cutoff = 350000) |> 
		plot(plot.numbers = TRUE)
variogram(NO2~1, no2.sf, width=5000, cutoff = 350000) |> 
		plot(plot.numbers = TRUE)
variogram(NO2~1, no2.sf, width=100000, cutoff = 350000) |> 
		plot(plot.numbers = TRUE)
variogram(NO2~1, no2.sf, width=1000, cutoff = 100000) |> 
		plot(plot.numbers = TRUE)
```

3. Fit a spherical model to the sample variogram of NO2, using `fit.variogram()` (follow the example below, replace "Exp" with "Sph")

```{r}
v = variogram(NO2~1, no2.sf)
v.fit = fit.variogram(v, vgm(1, "Sph", 50000))
```

4. Fit a Matern model ("Mat") to the sample variogram using different values
for kappa (e.g., 0.3 and 4), and plot the resulting models with the sample variogram.

```{r}
v.fit = fit.variogram(v, vgm(1, "Mat", 50000, kappa = .3))
plot(v, v.fit)
v.fit = fit.variogram(v, vgm(1, "Mat", 50000, kappa = 1))
plot(v, v.fit)
v.fit = fit.variogram(v, vgm(1, "Mat", 50000, kappa = 4))
plot(v, v.fit)
```

Note the different behaviour at the origin: curving at the origin
indicates (the assumption of) a very smooth process.

5. Which model do you like the best? Can the SSErr attribute of the
fitted model be used to compare the models? How else can variogram
model fits be compared?

Answer: hard to tell! A good fit to a sample variogram may not directly
mean a good performance when used for interpolation; a way to check the
latter is cross validation (function `krige.cv` in package `gstat`).

## Second round of exercises

1. What causes the differences between the mean and the variance of
the simulations (left) and the mean and variance obtained by kriging
(right)?

A: the number of realisations; the more simulations (realisations)
we would create and average over, the closer the result looks like
the kriging mean and variance.

2. When comparing (above) the sample variogram of simulated fields
with the variogram model used to simulate them, where do you
see differences? Can you explain why these are not identical, or
hypothesize under which circumstances these would become (more)
identical?

A: the differences are larger for larger distances. If we would
increase the size of the area over which we create simulations (say,
Europe rather than only Germany) then sample variogram values from
individual realisations would be closer to those of the variogram
model used to generate the simulations.

3. Under which practical data analysis problem would conditional
simulations be more useful than the kriging prediction + kriging
variance maps?

A. When uncertainties are computed and the kriging or simulation maps
are an input, and (i) part of the downstream computation process
is nonlinear (e.g.  we are interested in extremes, or exceedances
of a threshold) or (ii) spatial correlations are important, e.g.
because we do some kind of spatial aggregation.  The kriging map
values are always smoother, and exhibit less variability, then the
values interpolated.

