## Geostatistical data

### Learning goals

* Get familiar with geostatistical data and spatial interpolation
* Get familiar with concepts of geostatistics: stationarity, variogram, kriging, conditional simulation
* Get an idea what spatiotemporal geostatistics is about

### Reading materials

From [Spatial Data Science: with applications in R](https://r-spatial.org/book/): 

* Chapter 12: Spatial Interpolation
* Chapter 13: Multivariate and Spatiotemporal Geostatistics

### Exercises for Today

* Exercises of Ch 12: Spatial interpolation

::: {.callout-tip title="Summary"}

* Intro to `gstat`
* Geostatistical data
* Spatial correlation, variograms, stationarity
* Kriging
* Simulating geostatistical data
* Spatiotemporal geostatistics

:::

## `gstat`

R package `gstat` was written in 2002/3, from a stand-alone C
program that was released under the GPL in 1997. It implements
"basic" geostatistical functions for modelling spatial dependence
(variograms), kriging interpolation and conditional simulation.
It can be used for multivariable kriging (cokriging), as well as
for spatiotemporal variography and kriging. Recent updates included
support for `sf` and `stars` objects.

## What are geostatistical data?

Recall from day 1: locations + measured values

* The value of interest is measured at a set of sample locations
* At other location, this value exists but is _missing_
* The interest is in estimating (predicting) this missing value (interpolation)
* The actual sample locations are not of (primary) interest, the signal is in the measured values

```{r}
library(sf)
no2 <- read.csv(system.file("external/no2.csv",
    package = "gstat"))
crs <- st_crs("EPSG:32632") # a csv doesn't carry a CRS!
st_as_sf(no2, crs = "OGC:CRS84", coords =
    c("station_longitude_deg", "station_latitude_deg")) |>
    st_transform(crs) -> no2.sf
library(ggplot2)
# plot(st_geometry(no2.sf))
"https://github.com/edzer/sdsr/raw/main/data/de_nuts1.gpkg" |>
  read_sf() |>
  st_transform(crs) -> de
ggplot() + geom_sf(data = de) +
    geom_sf(data = no2.sf, mapping = aes(col = NO2))
```

## Spatial correlation

### Lagged scatterplots

"by hand", base R:
```{r}
(w = st_is_within_distance(no2.sf, no2.sf, units::set_units(50, km), 
						  retain_unique = TRUE))
d = as.data.frame(w)
x = no2.sf$NO2[d$row.id]
y = no2.sf$NO2[d$col.id]
cor(x, y)
plot(x, y, main = "lagged scatterplot")
abline(0, 1)
```

using gstat:
```{r}
library(gstat)
hscat(NO2~1, no2.sf, breaks = c(0,50,100,150,200,250)*1000)
```

### Variogram

"by hand", base R:
```{r}
z = no2.sf$NO2
z2 = 0.5 * outer(z, z, FUN = "-")^2
d = as.matrix(st_distance(no2.sf))
vcloud = data.frame(dist = as.vector(d), gamma = as.vector(z2))
vcloud = vcloud[vcloud$dist != 0,]
vcloud$dclass = cut(vcloud$dist, c(0, 50, 100, 150, 200, 250, 300, 350) * 1000)
v = aggregate(gamma~dclass, vcloud, mean)
plot(gamma ~ dclass, v, ylim = c(0, 20))
```

using gstat:
```{r}
vv = variogram(NO2~1, no2.sf, width = 50000, cutoff = 350000)
vv$gamma - v$gamma
plot(vv)
```

::: {.callout-note title="Breakout session"}

Compute the variogram of NO2 using argument `cloud = TRUE`. 

* how does the resulting object differ from the "regular" variogram 
* what do the "left" and "right" fields refer to?
* when we plot the resulting object, does it still indicate spatial correlation?

:::


##  Interpolation

For interpolation, we first need a target grid (point patterns have
an observation window, geostatistical data not!)

A simple interpolator (that is hard to beat) is the inverse distance interpolator,
$$\hat{Z}(s_0) = \sum_{j=1}^n \lambda_j Z(s_i)$$
with $\lambda_j$ proportional to $||s_i - s_0||^{-p}$ and normalized to sum
to one (weighted mean), and $p$ tunable but defaulting to 2.

Using the data range:
```{r}
library(stars)
g1 = st_as_stars(st_bbox(no2.sf))
library(gstat)
idw(NO2~1, no2.sf, g1) |> plot(reset = FALSE)
plot(st_geometry(no2.sf), add = TRUE, col = 'yellow')
plot(st_cast(st_geometry(de), "MULTILINESTRING"), add = TRUE, col = 'red')
```

Better to use the outer polygon:
```{r}
g2 = st_as_stars(st_bbox(de))
idw(NO2~1, no2.sf, g2) |> plot(reset = FALSE)
plot(st_geometry(no2.sf), add = TRUE, col = 'yellow')
plot(st_cast(st_geometry(de), "MULTILINESTRING"), add = TRUE, col = 'red')
```

And crop to (mask out outside) the area of interest:
```{r}
g3 = st_crop(g2, de)
i = idw(NO2~1, no2.sf, g3) 
plot(i, reset = FALSE, main = "yearly mean NO2, rural background")
plot(st_geometry(no2.sf), add = TRUE, col = 'yellow')
plot(st_cast(st_geometry(de), "MULTILINESTRING"), add = TRUE, col = 'red')
```

Geostatistical approaches compute weights based on covariances
between observations $Z(s_i)$, and between observations and the
value at the interpolation location $Z(s_0)$. These covariances
are obtained from a model fitted to the sample variogram.

## Fit a variogram model

```{r}
# The sample variogram:
v = variogram(NO2~1, no2.sf)
plot(v)
```

fit a model, e.g. an exponential model:
```{r}
v.fit = fit.variogram(v, vgm(1, "Exp", 50000))
plot(v, v.fit)
```

Given this model, we can interpolate using the _best unbiased linear
predictor_ (BLUP), also called kriging predictor:

```{r}
k = krige(NO2~1, no2.sf, g3, v.fit)
k$idw = i$var1.pred
k$kriging = k$var1.pred
hook = function() {
  plot(st_geometry(no2.sf), add = TRUE, col = 'yellow')
  plot(st_cast(st_geometry(de), "MULTILINESTRING"), add = TRUE, col = 'red')
}
plot(merge(k[c("kriging", "idw")]), hook = hook, breaks = "equal")
```

::: {.callout-important}
## Density or interpolation?

Both density maps shown in the Point Pattern section and interpolated
maps shown in this section look very similar:

* raster maps with continuous values
* smooth spatial patterns

The differences are huge!

* point density estimates estimate the number of points per unit area; the values are (normalized) _counts_
* interpolated maps estimate an unmeasured continuous variable; the values are weighted averages of an _attribute_

```{r echo=FALSE}
library(spatstat)
p = as.ppp(st_geometry(no2.sf), W = as.owin(st_union(st_geometry(de))))
density(p) |> st_as_stars() |> plot(main = "density: # / km^2", reset = FALSE)
plot(st_geometry(no2.sf), add = TRUE, col = 'yellow')
```
```{r echo=FALSE}
plot(i, main = "interpolation: NO2", reset = FALSE)
plot(st_geometry(no2.sf), add = TRUE, col = 'yellow')
```

:::

## Conditional simulation

### Simulating spatially correlated data

Using a coarse grid, with base R:
```{r}
set.seed(13579)
g2c = st_as_stars(st_bbox(de), dx = 20000)
g3c = st_crop(g2c, de)
p = st_as_sf(g3c, as_points = TRUE)
d = st_distance(p)
Sigma = variogramLine(v.fit, covariance = TRUE, dist_vector = d)
n = 100
ch = chol(Sigma)
sim = matrix(rnorm(n * nrow(ch)), nrow = n) %*% ch + mean(no2.sf$NO2)
for (i in seq_len(n)) {
	m = g3c[[1]]
	m[!is.na(m)] = sim[i,]
	g3c[[ paste0("sim", i) ]] = m
}
plot(merge(g3c[2:11]), breaks = "equal")
```

As a check, we could compute the variogram of some of the realisations:
```{r}
g3c["sim4"] |> 
  st_as_sf() |> 
  variogram(sim4~1, data = _) |> 
  plot(model = v.fit, ylim = c(0,16.5))
g3c["sim5"] |> 
  st_as_sf() |> 
  variogram(sim5~1, data = _) |> 
  plot(model = v.fit, ylim = c(0,16.5))
g3c["sim6"] |> 
  st_as_sf() |> 
  variogram(sim6~1, data = _) |> 
  plot(model = v.fit, ylim = c(0,16.5))
```

The mean of these simulations is constant, not related to measured values:
```{r}
st_apply(merge(g3c[-1]), c("x", "y"), mean) |> plot()
mean(no2.sf$NO2)
```

Conditioning simulations on measured values can be done with `gstat`, using
_conditional simulation_

```{r}
cs = krige(NO2~1, no2.sf, g3, v.fit, nsim = 50, nmax = 30)
plot(cs[,,,1:10])
```

We see that these simulations are much more alike;
```{r}
csm = st_apply(cs, c("x", "y"), mean)
csm$kriging = krige(NO2~1, no2.sf, g3, v.fit)[1]
plot(merge(csm), breaks = "equal")
csv = st_apply(cs, c("x", "y"), var)
csv$kr_var = krige(NO2~1, no2.sf, g3, v.fit)[2]
plot(merge(csv), breaks = "equal")
```

## SpatioTemporal geostatistics

* [Ch 13](https://r-spatial.org/book/13-Geostatistics.html)

## Further reading

* Pebesma, E.J., 2004. Multivariable geostatistics
in S: the gstat package. Computers & Geosciences, 30:
[683-691](https://doi.org/10.1016/j.cageo.2004.03.012).

* Benedikt Gräler, Edzer Pebesma and Gerard Heuvelink, 2016.
Spatio-Temporal Interpolation using gstat.  The R Journal 8(1),
[204-218](https://journal.r-project.org/archive/2016/RJ-2016-014/index.html)

