# Introduction {.unnumbered}

## Required packages

The following packages may be used during the course; it is assumed
that you know how to install packages, and have permission to do
so on your computer. 

### CRAN packages:

```{r eval=FALSE}
install.packages(c("classInt",
"colorspace",
"dplyr",
"ggplot2",
"gstat",
"hglm",
"igraph",
"lme4",
"lwgeom",
"maps" ,
"mapview",
"randomForest",
"rnaturalearth",
"s2",
"scales",
"sf",
"sp",
"spacetime",
"spdep",
"spatialreg",
"spatstat",
"spData",
"stars",
"terra",
"tidyverse",
"tmap",
"units",
"viridis",
"viridisLite",
"xts"))
```

### non-CRAN packages:

```{r eval=FALSE}
install.packages("spDataLarge", repos = "https://nowosad.github.io/drat/", 
				 type = "source")
install.packages("starsdata", repos = "http://cran.uni-muenster.de/pebesma/", 
				 type = "source")
```

### Introduction to the course

* introduction of the tutor
* introduction of course participants, please state
   * name,
   * where you're from,
   * what kind of spatial data analysis you have done so far

### How we work

* live sessions are from 15:00-18:00 CET (Berlin time)
* slack communication during the full week
* please share open questions you run into in your actual research, preferably with (example) data and R code

### Resources

* [Spatial Data Science: With applications in R](https://r-spatial.org/book/), by Pebesma and Bivand 2023
* Vignettes of [sf](https://r-spatial.github.io/sf/): tab "Articles"
* Vignettes of [stars](https://r-spatial.github.io/stars/): tab "Articles"
* All these material are written using R-markdown (or [quarto](https://quarto.org/))

### Why R?

* R is old! Think of the advantages!
* R is as good as any data science language, but is more in focus with the statistical community
* R has a strong ecosystem of users and developers, who communicate and collaborate (and compete, mostly in a good way)
* R spatial packages have gone full cycle: 
    * the first generation has been deprecated
    *  removed from CRAN, and 
    * replaced by modern versions
* R is a data science language that allows you [to work reproducibly](https://eartharxiv.org/repository/view/3216/)
* Because we have [CRAN](https://cran.r-project.org/)

### Reproducing the current course

* Go to https://github.com/edzer/sswr/
* Go to "Code", "copy URL to clipboard"
* Clone this repo to your hard drive
* Start one of the qmd files by double clicking, or on the command line with RStudio, or using some other tooling
* Run the code sections!
